Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-14 write-ups/PLDI/paper.tex
added some sample images
added one more figure
Shortening

\item The model is backwards-compatible: if $\offerkill$ and
  $\offerfork$ are ignored, a cooperative blocking kernel will behave
  exactly as a regular blocking kernel does on current GPUs.

\item The model can be implemented on top of the occupancy-bound
  execution model that current GPU platforms provide---we present a
  proof-of-concept implementation on top of OpenCL 2.0 that requires
  no special-purpose hardware or driver support.

@@

Placing the $\offerkill$ and $\offerfork$ primitives requires manual
effort and care, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
suggests that this is straightforward in practice.  We discuss this in
detail in \mysec\ref{sec:portingalgorithms}.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top of the OpenCL 2.0 programming model
  (\mysec\ref{sec:implementation}).

\item An experience report of porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  using three GPUs assessing the overhead associated with cooperative
  kernels, and the responsiveness of the multi-tasking that the
  approach enables (\mysec\ref{sec:experiments}).

@@

We provide background on the industry-standard OpenCL programming
model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
and illustrate OpenCL using two irregular data-parallel algorithms---a
work stealing queue and breadth-first graph search---that also
pinpoint the problems associated with unfair scheduling of workgroups
and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}).

@@

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is written in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may invoke the $\getglobalid()$
function to query its unique thread id in order to access different
data or follow different control flow paths to other threads.

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of threads.  Builtin functions
$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
}  The number
of threads per workgroup and the number of workgroups are obtained via
$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
id are related via the equation $\getglobalid() = \getgroupid() \times
\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize()$, which is equal to $\getlocalsize()\times\getnumgroups()$.

The execution of all threads in a workgroup can be synchronised via a
workgroup barrier, which also ensures memory consistency.  This
barrier is a \emph{workgroup-level} function: it must be encountered
uniformly by all or none of the threads of a workgroup, thus can only
appear in conditional code if all threads in a workgroup that reach
the barrier agree on the guards of all enclosing conditional
statements.  As discussed below, a \emph{global} barrier
(synchronising all threads running a kernel) is \emph{not} provided as
a primitive.

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
kernel.  Each workgroup has its own portion of \emph{local} memory,
which can be used for fast intra-workgroup communication.  Every thread has
an allocation of very fast \emph{private} memory, where
function-local variables are allocated.

Communication between threads in the same workgroup can be achieved
using the workgroup barrier construct described above.  Finer-grained
communication within a workgroup, as well as inter-workgroup
communication and communication with the host while the kernel is
running, is enabled by a set of atomic data types and operations.  In
particular, fine-grained host/device communication is via atomic
operations on SVM.

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups so
  a correct algorithm cannot assume that workgroups will execute in
  parallel.  There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} OpenCL is not alone in this
regard: CUDA similarly provides no guarantees~\cite{...}.

@@

>>>>

\item The model is backwards-compatible: if $\offerkill$/$\offerfork$ are ignored, a cooperative kernel will behave
  exactly as a regular kernel does on current GPUs.

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide---we present a
  proof-of-concept implementation that requires
  no special-purpose hardware or driver support.

@@

Placing the new primitives requires manual effort, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
(\mysec\ref{sec:portingalgorithms}) suggests that this is straightforward in practice.

@@

\item \emph{Cooperative kernels}, an extended GPU programming model that support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}).

\item Experimental results assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).

@@

We outline the industry-standard OpenCL programming model on which we
base cooperative kernels (\mysec\ref{sec:opencl}), and illustrate
OpenCL and the scheduling requirements of irregular algorithms using two examples: a work stealing queue and frontier-based graph traversal
(\mysec\ref{sec:openclexamples}).

@@

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is written in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments.  A thread can call $\getglobalid()$
to obtain its unique id, to access distinct data or follow different control flow paths to other threads.

The threads executing a kernel are divided into a number of \emph{workgroups}, each consisting of a fixed number of threads.  Builtin functions
$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
its workgroup and the id of the workgroup, respectively.\footnote{For simplicity we restrict attention to one-dimensional kernels, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
}  The number
of threads per workgroup and the number of workgroups are obtained via
$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
id are related via the equation $\getglobalid() = \getgroupid() \times
\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize() = \getlocalsize()\times\getnumgroups()$.

Execution of the threads in a workgroup can be synchronised via a
workgroup barrier, which also ensures memory consistency.  This
barrier is a \emph{workgroup-level} function: it can only
appear in conditional code if all threads in a workgroup that reach
the barrier agree on the guards of all enclosing conditional
statements.  A \emph{global} barrier
(synchronising all threads of a kernel) is \emph{not} provided as
a primitive.

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
kernel.  Each workgroup has its own portion of \emph{local} memory for fast intra-workgroup communication.  Every thread has
an allocation of very fast \emph{private} memory, where
function-local variables are allocated.

Communication within a workgroup can be achieved
using a workgroup barrier.  Finer-grained
communication within a workgroup, as well as inter-workgroup
communication and communication with the host while the kernel is
running, is enabled by a set of atomic data types and operations.  In
particular, fine-grained host/device communication is via atomic
operations on SVM.

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups 
%so
%  a correct algorithm cannot assume that workgroups will execute in
%  parallel.  
\dots There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} CUDA similarly provides no guarantees~\cite{...}.

@@

\begin{figure*}
\includegraphics[width=.7\columnwidth]{iris_octree_NA.pdf} 
\includegraphics[width=.7\columnwidth]{hd520_mis_ecology.pdf}
\includegraphics[width=.7\columnwidth]{hd5500_bfs_rmat22.pdf}
\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
\end{figure*}
